
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# MB5370: Techniques in Marine Science 1

# Module 2 - Workshop 02. Advanced data wrangling


Load packages

```{r}
library(tidyverse)
library(palmerpenguins)
```

### Tiding data example

```{r}
table1
```

```{r}
table2
```

```{r}
table3
```

Compute rate per 10,000

```{r}
table1 %>%
  mutate(rate = cases / population * 10000)
```

Compute cases per year

```{r}
table1 %>%
  count(year, wt = cases)
```

Create plot to visualise changes over time

```{r}
ggplot(table1, aes(year, cases)) +
  geom_line(aes(group = country), colour = "grey50") +
  geom_point(aes(colour = country))
```

Lenghting dataset example \####

```{r}
billboard

billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank"
  )

billboard |>
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    values_drop_na = TRUE
  )
```

Pivoting longer - example \####

```{r}
df <- tribble(
    ~id,  ~bp1, ~bp2,
    "A",  100,  120,
    "B",  140,  115,
    "C",  120,  125
  )
```

Make dataset have 3 variables (id, measurement, value)

```{r}
df |> 
  pivot_longer(
    cols = bp1:bp2,
    names_to = "measurement",
    values_to = "value"
  )
```

### Widening dataset exercise

```{r}
cms_patient_experience
```

See complete sets of values for measure_cd and measure_title by using
distinct()

```{r}
cms_patient_experience |>
  distinct(measure_cd, measure_title)
```

Pivot wider

```{r}
cms_patient_experience |>
  pivot_wider(
    names_from = measure_cd,
    values_from = prf_rate
  )
```

Tell pivot_wider() which coloumns have values that uniquely identify
each row: "org"

```{r}
cms_patient_experience |>
  pivot_wider(
    id_cols = starts_with("org"),
    names_from = measure_cd,
    values_from = prf_rate
  )
```

### Pivoting wider - example

```{r}
df <- tribble(
  ~id, ~measurement, ~value,
  "A",        "bp1",    100,
  "B",        "bp1",    140,
  "B",        "bp2",    115, 
  "A",        "bp2",    120,
  "A",        "bp3",    105
)
```

Take the names from the measurement column using the names_from()
argument and the values from the value column using the values_from()
argument

```{r}
df |> 
  pivot_wider(
    names_from = measurement,
    values_from = value
  )

df |> 
  distinct(measurement) |> 
  pull()

df |> 
  select(-measurement, -value) |> 
  distinct()

df |> 
  select(-measurement, -value) |> 
  distinct() |> 
  mutate(x = NA, y = NA, z = NA)
```

### Create long version of the penguins dataset exercise

```{r}
penguins_long <- penguins |>
  pivot_longer(
    cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g),
    names_to = "measurement_type",
    values_to = "value"
  )
```

View the result

```{r}
head(penguins_long)
```

Plot data

```{r}
penguins_long |>
  drop_na(value) |>
  ggplot(aes(x = value, fill = species)) +
  geom_histogram(bins = 30, alpha = 0.7, colour = "black") +
  facet_wrap(~ measurement_type, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Morphometric distributions across penguin species",
    x = "Measurement value",
    y = "Frequency"
  )
```

Widening data with pivot_wider()

```{r}
mass_summary <- penguins |>
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(mean_mass = mean(body_mass_g))
```

```{r}
head(mass_summary)
```

Create clean matrix of species by island

```{r}
mass_matrix <- mass_summary |>
  pivot_wider(
    names_from = island,
    values_from = mean_mass
  )
```

```{r}
head(mass_matrix)
```

### Separating and uniting data tables

```{r}
table3
```

Split rate coloumn into two variables (1: cases, 2: population) using
separate()

```{r}
table3 %>%
  separate(rate, into = c("cases", "population"))
```

```{r}
table3 %>%
  separate(rate, into = c("cases", "population"), sep = "/")
```

Change cases and population from characters to interger

```{r}
table3 %>%
  separate(rate, into = c("cases", "population"), convert = TRUE)
```

Separate the last two digits of each year

```{r}
table3 %>%
  separate(year, into = c("century", "year"), sep = 2)
```

Use unite() to inverse separate() and rejoin century and year columns

```{r}
table5 %>%
  unite(new, century, year, sep = "")
```

### Wrangling strings and dates

A remarkably messy data frame of field sites

```{r}
messy_sites <- tibble(
  site_id = c(" Nelly Bay", "nelly_bay", "NELLY BAY", " Geoffrey_Bay ", "geoffrey bay")
)
```

Using stringr within mutate to standardize the text

```{r}
clean_sites <- messy_sites |>
  mutate(
    # 1. Convert everything to lowercase
    site_clean = str_to_lower(site_id),
    # 2. Trim any leading or trailing whitespace (invisible spaces at the ends)
    site_clean = str_trim(site_clean),
    # 3. Replace any spaces with underscores
    site_clean = str_replace_all(site_clean, pattern = " ", replacement = "_")
    
  )
```

```{r}
print(clean_sites)
```

Load packages

```{r}
library(lubridate)
```

Parsing different date formats

```{r}
date_1 <- dmy("25/12/2026")
date_2 <- ymd("2026-12-25")
```

R now recognizes these as identical Date objects

```{r}
date_1 == date_2
```

A tibble of raw sensor data with a messy character timestamp

```{r}
sensor_data <- tibble(
  raw_time = c("14-05-2026 08:30:00", "14-05-2026 08:45:00", "14-05-2026 09:00:00"),
  temperature = c(24.5, 24.6, 24.4)
)
```

Converting character strings to true POSIXct datetime objects

```{r}
sensor_clean <- sensor_data |>
  mutate(
    true_time = dmy_hms(raw_time)
  )
```

```{r}
print(sensor_clean)
```

### Relational data: joining tables

Table 1: Biological observation data

```{r}
observations <- tibble(
  site_code = c("NB", "GB", "MI", "NB"),
  species = c("Trout", "Snapper", "Trout", "Cod"),
  count = c(5, 2, 1, 3)
)
```

Table 2: Spatial metadata

```{r}
site_metadata <- tibble(
  site_code = c("NB", "GB", "MI", "RP"),
  zone = c("Marine National Park", "Conservation Park", "Habitat Protection", "General Use"),
  lat = c(-19.16, -19.15, -19.14, -19.12)
)
```

Workhorse: left_join() Most common and safest Joining metadata to our
observations

```{r}
joined_data <- observations |>
  left_join(site_metadata, by = join_by(site_code))
```

```{r}
print(joined_data)
```

The strict filter: inner_join() Keeps rows that have complete matches in
both tables

```{r}
matched_data <- observations |>
  inner_join(site_metadata, by = join_by(site_code))
glimpse(matched_data)
```

The detective: anti_join() Does not add new columns instead it filters
your left table to only show rows that do not have a match in the right
table

```{r}
missing_context <- observations |>
  anti_join(site_metadata, by = join_by(site_code))
```

### Handling missing values

Converting legacy errors with na_if()

Raw data from an old temperature logger

```{r}
logger_data <- tibble(
  depth_m = c(10, 20, 30, 40),
  temp_c = c(24.5, 24.1, -999, 23.5)) # -999 is a known sensor error code
```

Convert the -999 error codes to true NA values

```{r}
fixed_logger <- logger_data |>
  mutate(temp_c = na_if(temp_c, -999))
```

```{r}
print(fixed_logger)
```

Replacing missing values with coalesce()

Simulate some count data

```{r}
shark_counts <- tibble(
  site = c("Reef_A", "Reef_B", "Reef_C"),
  shark_count = c(3, NA, 5)) # The NA here actually means 0 sharks were seen
```

Replace NA with 0

```{r}
shark_fixed <- shark_counts |>
  mutate(shark_count = coalesce(shark_count, 0))
```

```{r}
print(shark_fixed)
```

The special case of NaN (not a number)

```{r}
cpue_data <- tibble(
  site = c("Bay_1", "Bay_2"),
  catch = c(10, 0),
  effort_hours = c(2, 0))
```

Calculating CPUE (catch / effort)

```{r}
cpue_calc <- cpue_data |>
  mutate(
    cpue = catch / effort_hours)
```

```{r}
print(cpue_calc)
```

Implicit missing values and the zero-data framework: complete()

```{r}
raw_catch <- tibble(
  site = c("Reef_1", "Reef_1", "Reef_2"),
  species = c("Pmaculatus", "Pleopardus", "Pmaculatus"),
  count = c(5, 2, 8))
```

Force inclusion of hidden zero-catch data

```{r}
full_catch_matrix <- raw_catch |>
  complete(site, species, fill = list(count = 0))
```

```{r}
print(full_catch_matrix)
```

The nuclear option: drop_na()

```{r}
sensor_log <- tibble(
  day = 1:4,
  salinity = c(35.2, 35.1, NA, 35.3)
)
```

Remove any row where salinity is missing

```{r}
clean_log <- sensor_log |>
  drop_na(salinity)
```

### Penguins exercise 1

Load packages

```{r}
library(tidyverse)
library(lubridate)
library(palmerpenguins)
```

Messy data

```{r}
island_metadata <- tibble(
  island_name = c(" biscoe", "Dream ", "Torgersen"),
  station_install = c("15/01/2003", "22-03-2004", "05/11/2001"),
  latitude = c(-64.81, -64.73, -64.76)
)
```

```{r}
print(island_metadata)
```

### Clean data

Fix the capital letters Fix the dates

```{r}
clean_metadata <- island_metadata |>
  mutate(
    island_name = str_squish(island_name) |> str_to_title(),
    station_install = parse_date_time(station_install, orders = "dmy") |> as.Date()
  )
```

```{r}
print(clean_metadata)
```

### Penguins exercise 2

Merge clean dataset with biological data Change island_name to island

```{r}
penguins_spatial <- penguins |>
  left_join(clean_metadata, by = join_by(island == island_name))
```

Check that it worked

```{r}
head(penguins_spatial)
```

### Penguins exercise 3

Drop missing values in the body_mass_g coloumn Group by species and
island Summarise the data to find the maximum body mass Pivot dataset
wider so that species forms the rows and islands form the columns

```{r}
wide_summary <- penguins_spatial |>
  drop_na(body_mass_g) |>
  group_by(species, island) |>
  summarise(max_body_mass = max(body_mass_g), .groups = "drop") |>
  pivot_wider(names_from = island, values_from = max_body_mass)
```

```{r}
print(wide_summary)
```

Still have NA values because some of the penguin species are not found
on those islands.
